---
title: "Untitled"
author: "Nishat Nini Urmi"
date: "6/3/2020"
output: html_document
editor_options: 
  chunk_output_type: inline
---
 
##Part 1: Importing files, data wrangling, mathematical operations, plots and saving code on GitHub 

### Question 1 : Read in the file, making the gene accession numbers the row names. Show a table of values for the first six genes
####Answer: The "geneexpression.tsv"was downloaded from the link "https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/gene_expression.tsv", and saved into Rstudio in a file. 
```{r}
x<- read.table("data/gene_expression.tsv", header = TRUE, stringsAsFactors = FALSE, row.names = 1)
head(x)
str(x)
x[1:6,1:2]
```
####The row names were changed into gene accession number while naming the table as x
### Question 2 : Make a new column which is the mean of the other columns. Show a table of values for the first six genes
```{r}
x$Mean_cal <- rowMeans(x)
head(x)
x[1:6,1:3]
```
####A new column "Mean_cal" is created that holds the mean value of the other columns. 

###Question3. List the 10 genes with the highest mean expression 
```{r}
Highest_mean <-x[order(-x$Mean_cal),]
head(Highest_mean,10)
```
####The function "Highest mean" calculates the highest expression from the target column. 

###Question 4.Determine the number of genes with a mean <10 
```{r}
subset(x,Mean_cal<10)
```
####The subset function here shows the genes with only the mean values less than 10

###Question 5. Make a histogram plot of the mean values in png format and paste it into your report
```{r,"Histogram for Mean values",echo=FALSE}
hist(x$Mean_cal,main="Histogram for Mean values",xlab = "mean value",border = "red",col="blue",breaks = 50,xlim=c(0,10000))
```
###Question 6. Import this csv file into an R object. What are the column names?
####The CSV files were uploaded from the link "https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/growth_data.csv" and saved in Rstudio. The "csv" file was saved as obejct "y". 
```{r}
y <- read.csv("data/growth_data.csv", header = TRUE, stringsAsFactors = FALSE)
head(y)
str(y)
```
###Name of the columns are- "site", "TreeID","Circumf_2004_cm", "Circumf_2009_cm", "Circumf_2014_cm " and "Circumf_2019_cm"

###Question 7 Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites. 
#### Prior to calculating mean and Standard deviation, both sites of study were saved as seperate obejcts "NE" for "NorthEast" and "SE" for "SouthWest" 
```{r}

NE <- subset(y,Site=="northeast")
SW <- subset(y,Site=="southwest")
head(NE)
head(SW)
```

###Mean calculation of tree circumference at the start and end of the study for North East Site
```{r}
mean(NE$Circumf_2004_cm)
mean(NE$Circumf_2019_cm)
sd(NE$Circumf_2004_cm)
sd(NE$Circumf_2019_cm)
```
####Mean  and standard deviation calculation of tree circumfurence at the satrt and end of the site Southwest
```{r}
mean(SW$Circumf_2004_cm)
mean(SW$Circumf_2019_cm)
sd(SW$Circumf_2004_cm)
sd(SW$Circumf_2019_cm)
```

